Generative AI for image reconstruction in Intensity Interferometry: a first attempt

### 2025-09-30 Phase_II/draft/image-reconstruction-past-efforts.tex
Section Introduction re-written and anew section on Past efforts introduced

\section{Introduction}

Intensity Interferometry (II) was first reported by Hanbury Brown and Twiss (HBT) during the 1950s \citep{brown1954lxxiv, HBT56} as a \textquotedblleft new type of interferometry\textquotedblright\ to measure stellar parameters such as angular diameter, orbits, and limb darkening coefficients. Later, theoretical results reported by \cite{brown1957interferometry, brown1958interferometry}, along with those of \cite{glauber1963quantum} and others, demonstrated the deeper physical properties of photon correlations that lie at the core of II and laid the foundation for Quantum Optics \cite[for textbook treatments see][]{MandelWolf1995, Hecht2002}.

By the 1970s, with stellar parameter measurements of 32 stars in single and multiple star systems conducted by Hanbury Brown and his collaborators \citep{hanbury1974angular} at the historic Narrabri Stellar Intensity Interferometer (NSII) in Australia, II had emerged as an alternative to the already established technique of Michelson Interferometry for measuring stellar parameters. Despite these significant achievements, the method did not gain widespread adoption in the ensuing decades, primarily due to the unavailability of sensitive photon detectors and advanced data analysis equipment.

More recently, proposals to utilize Imaging Atmospheric Cherenkov Telescope (IACT) facilities for conducting II observations of stars have emerged\citep{LeBohec2006, nunez2010stellar, nunez2012high, 2013APh....43..331D}. It has been demonstrated that such observations could be carried out during bright, moonlit nights when $\gamma$-ray observations based on upper atmospheric Cherenkov showers were not feasible. This approach has the potential to enhance the scientific output of existing IACT facilities, and especially of the upcoming Cherenkov Telescope Array Observatory (CTAO).  SII observations at VERITAS, MAGIC, and HESS are now being reported \citep[e.g.,][]{2024ApJ...966...28A,2024MNRAS.529.4387A,2025MNRAS.537.2334V}. Simulations \citep[e.g.,][]{10.1093/mnras/stab2391, 10.1093/mnras/stac2433} have argued that recent advancements in photon detectors could be effective in achieving high-precision measurements of parameters for stellar objects.

Beyond measuring stellar diameters and other parameters of star systems, a fundamental goal of optical astronomy is to image stellar systems at high angular resolution.  In the context of II, this involves reconstructing the source's image from the intensity correlations recorded by pairs of telescopes (light buckets) on the ground. However, because the primary observable in II is the intensity of electromagnetic field rather than the amplitude, the phase of the interferometric signal is lost. So, there is a challenge to recover the phase of the signal, since a complete reconstruction of a source's brightness distribution requires phase information.

Several theoretical and computational approaches for phase reconstruction with II have been proposed. \cite{gamo1963triple} introduced the concept of triple-intensity correlation, which \cite{goldberger1963use} subsequently applied in an experiment to observe scattered particles in microscopic systems. Sato conducted experiments to measure the diameter and phase of asymmetrical objects, suggesting that triple correlation could extend II to image stellar bodies \citep{sato1978imaging, sato1979computer, sato1981adaptive}. However, achieving a satisfactory signal-to-noise ratio (SNR) remained a significant challenge for this approach.

@@

In this paper, we propose a conditional Generative Adversarial Network (cGAN) model \citep[following][]{isola2017image} to reconstruct images of fast-rotating stars using their simulated Intensity Interferograms and simulated sky-intensity distributions as input data for training, testing, and validation. We consider four Imaging Cherenkov Telescope Arrays (IACTs) and simulate observation of a fast-rotating star. The image predicted by the trained GAN shows promising results in reconstructing the star’s shape and size. The reconstructed brightness distributions are then assessed using moments.

This paper is organized as follows. The next section discusses Intensity Interferometry, focusing on its signal and noise characteristics for fast-rotating stars along the Earth’s rotation. The following section introduces the GAN formulation and its structure. The fourth section details the parameter selection for training the GAN for image reconstruction. The fifth section presents the results of the trained GAN both visually and via image moments. Finally, the paper concludes with a discussion of the overall results.
\begin{figure}
	\centering
	\includegraphics[width=\linewidth]{fig/telescope.png}
	\caption{The telescope configuration with similar properties each used to simulate the signal for II observation.}
	\label{fig:teles}
\end{figure}

>>>>

\section{Past Efforts at Image Reconstruction in Intensity Interferometry}
Several theoretical and computational approaches for phase reconstruction with II have been proposed. \cite{gamo1963triple} introduced the concept of triple-intensity correlation, which \cite{goldberger1963use} subsequently applied in an experiment to observe scattered particles in microscopic systems. Sato conducted experiments to measure the diameter and phase of asymmetrical objects, suggesting that triple correlation could extend II to image stellar bodies \citep{sato1978imaging, sato1979computer, sato1981adaptive}. However, achieving a satisfactory signal-to-noise ratio (SNR) remained a significant challenge for this approach.

@@

Let us now discuss the details of our procedure and method.

\begin{figure}
	\centering
	\includegraphics[width=\linewidth]{fig/telescope.png}
	\caption{The telescope configuration with similar properties each used to simulate the signal for II observation.}
	\label{fig:teles}
\end{figure}
